---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Personal activity monitoring
============================


### Downloading and unzipping the data
Lets see if its downloaded already, if not lets do so.
```{r, cache =TRUE}
library(ggplot2)

destfile="./data.zip"
if(!file.exists(destfile)){
      url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
      download.file(url, destfile = "./data.zip")
}
unzip("./data.zip")


```
###data processing

Now we read the data to a data frame. then we summarize the data with summary, its string representation and the header.
```{r, cache=F}
data <- read.table(file = "activity.csv", header = TRUE, sep = ",", nrows =  17568, na.strings = NA)
summary(data)
str(data)
head(data)
```
 
now we transform the date to the YYYY-MM-DD format
```{r, cache = F}
data$date <- as.Date(data$date, "%Y-%m-%d")
class(data$date)
head (data$date)
```

## Assignment questions
### 1. What is mean total number of steps taken per day?

1. calculate the number of steps per day
```{r, cache=F}
steps_per_day <- with(data, tapply(steps, date, sum, na.rm = TRUE))
str (steps_per_day)
head (steps_per_day)
```

Now we make a histogram and print out the mean and median of the total number of steps taken each day.
```{r, cache = F}

hist(steps_per_day, xlab = "Total steps per day", col = "red", main = "Total steps taken each day (ignore missing data)")


mean_steps_per_day<-mean(steps_per_day)
median_steps_per_day<-median(steps_per_day)
mean_steps_per_day
median_steps_per_day
```
### 2. What is the average daily activity pattern?

First, we need to calculate the average number of steps taken per each 5-minutes interval (averaged across all days). Then we make a plot of intervals and the average numbers of steps taken.
```{r, cache = F}
steps_per_interval =  with(data, tapply(steps, interval, mean, na.rm = TRUE))
plot(unique(data$interval), steps_per_interval, type = "l", xlab = "5-minute interval", ylab = "steps per interval averaged over all days", main = "Average daily activity")
```

The 5 minute interval which on average contains the most steps
```{r, cache = F}
names (which(steps_per_interval==max(steps_per_interval)))
```

### 3. Inputing missing values 
1. report the number of NAs
```{r, cache=F}
sum(is.na(data))
```

Well change the missing values with the mean of the number of steps for that 5 minute interval, and place it into a copy of the original data called data2
```{r, cache=F}
data2 <- data
data2$steps_per_interval =  with(data2, tapply(steps, interval, mean, na.rm = TRUE))
data2[is.na(data2$steps),]$steps <- data2[is.na(data2$steps),]$steps_per_interval
head(data2)
```
We will now redraw the histogram to see how changing the NAs into mean values and see how it impacts the graph.

```{r, cache = F}
steps_per_day_2 <- with(data2, tapply(steps, date, sum, na.rm = T))
hist(steps_per_day_2, xlab = "Total number of steps per day", col = "wheat2", main = "Total number of steps taken each day (with filled missing data)")
mean(steps_per_day_2)
median(steps_per_day_2)
```
we can see the histogram has been skewed towards the mean of the number of steps now, instead of being skewed to zero steps

### 4. Are there differences in activity patterns between business days and weekends?

Add a new "weekday" column:

```{r}
data2$weekday = weekdays(strptime(data2$date,"%Y-%m-%d"))
```

subset the data2 data frame into weekend and weekday

```{r}
data_weekdays <- subset(subset(data2,weekday != 'Saturday'),weekday != 'Sunday')
data_weekends <- rbind(subset(data2,weekday == 'Saturday'),subset(data2,weekday == 'Sunday'))
```

Make a panel plot:

```{r}
par(mfrow=c(2,1))
with(aggregate(steps ~ interval, data_weekends, mean),plot(interval,steps,type="l",main="Weekend"))
with(aggregate(steps ~ interval, data_weekdays, mean),plot(interval,steps,type="l",main="Weekday"))
     
```
